A state space model reveals the rich and varied dynamics of replay

### 2020-07-10 Article.tex

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\DeclareMathOperator*{\E}{\mathbb{E}}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Moreover, there is now substantial evidence that a subset of these events express stationary representations, which are often excluded. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

 Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used fixed time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches (refs) and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work. We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. Our method provides a useful and powerful tool for understanding the dynamics of mental representations, and could be adapted for use in many brain areas.

@@

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state (see methods). With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time and the probability of position over time, irrespective of dynamic (Figure \ref{1}F, left and right plot respectively).

An examination of these two summaries reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and the probability of the stationary dynamic is high at the beginning of the simulated replay. The change in the character of the data then drives a rapid transition to the continuous dynamic, reflecting the trajectory-like spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, and as a result, the fragmented dynamic dominates for the last third. Importantly, these dynamics also inform the estimate of latent position of the animal, and the correct assignment of dynamic reduces the variance of the latent position estimate.

@@

After testing our model on a single SWR, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see methods) to ensure that we were included both the rare large events as well as the much more common small events.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of events contained spatially coherent content in any form, defined as containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see methods). We found that 22478 of 23382 or 96\% of SWRs had some spatially coherent structure while 3353 of 23382 or 14\% had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of  23382 or 19\% of SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent but not continuous content.

Our model is specified in the context of a marked point process associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

@@

Our results indicate that most events with coherent spatial content are not continuous. We therefore examined these events in more details. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, 
than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events  represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events  that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify "non-local stationary trajectories" and found these trajectories in 1585 of 23382 (~7\%) of classified SWRs. About half of these (731)  were stationary throughout the duration of the SWR. These non-local stationary trajectories most commonly represented  reward wells or choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}. 
Finally, we examined the overall activity levels across dynamics and found similar total spiking rates, indicating ...

 (Figure \ref{5}E.

@@

Why our results are different from the Bayesian decoder
seek to describe all SWR events, clusterless decoding, more permissive, descriptive model

@@

model represents middle ground between HMM and typical Bayesian decoder. allows SWR to quickly be identified because of the discrete states. Can extend to incorporate other previously experienced environments. Can work in 2D. Can precisely identify timing of dynamics because of acausal decoding

adjudicates/clarifies results of Pfiefer and Foster and Stella et al.

relationship to Johnson and Redish 2008

extension to previously experienced tracks

@@

\subsection*{Simulated Data}
Encoding data for Figure \ref{1} and Figure \ref{fig:Figure1-Figure supplement 1} was generated by simulating 15 traversals of a 180 cm linear track. The spiking of 19 neurons was simulated using an inhomogeneous Poisson process where the instantaneous firing rate of each neuron changes according to a Gaussian place field. The Gaussian place fields had a variance of 36 cm and were spaced at 10 cm intervals over the 180 cm track. The decoding data for Figure \ref{1}G was generated by simulating 20,000 linear traversals of the 180 cm track, each with a unique constant speed, starting at 0.5 cm/s and increasing by 0.5 cm/s up to 10,000 cm/s. Each neuron spiked when the traversal passed through the peak location of its place field.

\subsection*{Recording Locations and Techniques}
Ten male Long Evans rats (500-700 g, 4-9 months old) were trained on a W-track spatial alternation task. 9 rats contributed to previous studies \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020}. Neural activity was recorded in CA1, CA2, CA3, MEC, Subiculum, and DG depending on the rat. We only used hippocampal tetrodes (CA1, CA2, CA3) in this study.

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 15 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals alternated between two types of runs: inbound and outbound. On inbound runs, the animal had to run from an outer arm to the center arm. On outbound runs, the animals had to run from the center arm to the least recently visited outer arm. Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from video of 2 LEDs placed on the head stage. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track, where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

\subsection*{SWR Detection}
Sharp wave ripples were detected using the same method as in \cite{Kayhippocampalnetworkspatial2016}. Briefly, each CA1 LFP was filtered between 150-250 Hz, squared and then summed across tetrodes---forming a single population trace over time. This trace was smoothed with a Gaussian with a 4 ms standard deviation and the square root of this trace was taken to get an estimate of the population ripple band power. Candidate SWR times were found by z-scoring the population power trace and finding times when the z-score exceeded 2 standard deviations for a minimum of 15 ms and the speed of the animal was less than 4 cm/s. The SWR times were then defined as times when the z-score was greater than the mean and contained the candidate SWR times. The code used for ripple detection can be found at \url{https://github.com/Eden-Kramer-Lab/ripple_detection}.

\subsection*{The Model}
Let $x_{k}$ be a continuous latent variable that corresponds to the position represented by the population of cells at time $t_k$ and $I_{k}$ be a discrete latent variable that is an indicator for the dynamics we wish to characterize: stationary, continuous, and fragmented. The goal of the model is to simultaneously estimate the posterior probability of both variables $p(x_k, I_k \mid O_{1:T})$, where $O_{1:T}$ corresponds to the observed data for for all times from 1 to $T$, the final time. The observed data can be either spike trains $\Delta N_{1:T}^{(1:C)}$ from $C$ putative cells when decoding with sorted spikes or multiunit spikes $\Delta N_{1:T}^{(1:E)}$ and their associated wave form features $\Vec{m}$ from each tetrode $E$ when decoding with clusterless spikes.

We have previously shown [Asilomar REF] that the posterior $p(x_k, I_k \mid O_{1:T})$ can be directly estimated by applying the following recursive causal filter equation, starting with initial conditions $p(x_{0}, I_{0} \mid O_{0})$:

$$
p(x_{k}, I_{k} \mid O_{1:k}) \propto
p(O_{k}  \mid x_{k}, I_{k}) \sum_{I_{k-1}} \int p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) 
Pr(I_{k} \mid I_{k-1}) p(x_{k-1}, I_{k-1} \mid O_{1:k-1}) dx_{k-1}
$$
and then applying the acausal smoother equation, starting from the last estimate of the casual filter $p(x_{T}, I_{T} \mid O_{1:k})$ and iterating backwards to time 1:
$$
p(x_{k}, I_{k} \mid O_{1:T}) =
p(x_{k}, I_{k} \mid O_{1:k})
\sum_{I_{k+1}} \int \frac{p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k})
Pr(I_{k+1} \mid I_{k})}{p(x_{k+1}, I_{k+1} \mid O_{1:k})}
p(x_{k+1}, I_{k+1} \mid O_{1:T}) dx_{k+1}
$$
where:
$$
p(x_{k+1}, I_{k+1} \mid O_{1:k}) =
\sum_{I_{k}} \int p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k}) Pr(I_{k+1} \mid I_{k})
p(x_{k}, I_{k} \mid O_{1:k}) dx_{k}
$$

@@

We set the dynamics transition matrix $Pr(I_{k} \mid I_{k-1})$, which defines how likely the dynamic is to change to another dynamic versus persist in the same dynamic, to be:

@@

We set the dynamics movement model $p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1})$, which defines how likely the position $x_{k}$ is to change given the previous position $x_{k-1}$ and current $I_{k}$ and past dynamics $I_{k-1}$, to be:

\begin{equation*}
  p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) = 
  \begin{blockarray}{*{3}{c} l}
    \begin{block}{*{3}{>{$\footnotesize}c<{$}} l}
      $I_{k} = Stationary$ & $I_{k} = Continuous$ & $I_{k} = Fragmented$ \\
    \end{block}
    \begin{block}{[*{3}{c}]>{$\footnotesize}l<{$}}
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0) & \mathcal{U}(\min x, \max x) \bigstrut[t]& $I_{k-1} = Stationary$ \\
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0)  & \mathcal{U}(\min x, \max x) & $I_{k-1} = Continuous$ \\
      \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & $I_{k-1} = Fragmented$ \\
    \end{block}
  \end{blockarray}
\end{equation*}
where $\delta(x_{k-1})$ is an identity transition matrix where position cannot change from the previous time step, $\mathcal{N}(x_{k-1}, 6.0)$ is a random walk from the previous position with variance 6.0, and $\mathcal{U}(\min x, \max x)$ is a uniform transition that allows transitions to any possible position.

As discussed in the Results, this says that when persisting in the same dynamic, the stationary, continuous, and fragmented dynamics are defined by the identity transition, the random walk, and the uniform transition, respectively. When transitioning to or from the fragmented dynamic, we assume we do not have any information about the position, so the transition is uniform. Finally, when the transition is from the stationary to continuous, we assume the position is spatially close where it was previously, so we use a random walk. When the transition is from continuous to stationary, we assume that the position is no longer changing, so we use the identity transition.

>>>>

\documentclass[times, twoside]{zHenriquesLab-StyleBioRxiv}
\usepackage{blindtext}
\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{mathtools}
\usepackage{blkarray, bigstrut}
\usepackage{bbold}
\usepackage{amsmath}
\DeclareMathOperator*{\E}{\mathbb{E}}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\norm}[1]{\left\lVert #1 \right\rVert}

@@

While this approach identifies constant speed events, it does not consider events that are rejected by the statistical test, and the use of a fixed size temporal bin acts as a boxcar smoother that limits the potential movement speeds of the representation. The linear fit is also problematic, because it has the potential to reject real events that do not move at constant speeds. Moreover, there is now evidence that a subset of these events have stationary representations of a single location, which are often excluded. In addition, these methods identify only a subset of events as containing content (typically somewhere between 10 and 45\% of the chosen events), and thereby provide little insight into the remaining majority of the events.

 Recognizing the problems with the linear fit, several studies have moved away from the constant velocity assumption, using the weighted mean or maximum of each posterior time bin and connecting each time bin with a line. For example, using this approach, Pfeiffer and Foster \cite{PfeifferAutoassociativedynamicsgeneration2015} found that replays can alternate between representing a single location and sequential spatial trajectories. On the other hand, Stella et al. \cite{StellaHippocampalReactivationRandom2019} found that replays are more spatially continuous, following Brownian diffusion dynamics. Both methods still used large time bins and neither took into account the uncertainty of the decoded estimates, making it hard to identify the source of the different conclusions.

@@

We therefore developed a state space model of generative activity that achieves all of those goals. State space models are a well-understood, well-known statistical solution to the problems described above. By mathematically modeling the relationship between the data and latent dynamics, state space models make the assumptions of the model explicit and interpretable. Our model goes beyond previous approaches \cite{MaboudiUncoveringtemporalstructure2018, DengRapidclassificationhippocampal2016} and characterizes represented trajectories as a mixture of three underlying patterns of dynamics: stationary trajectories, continuous trajectories that are many times the typical speed of the animal, and spatially fragmented trajectories. We show how this model can take advantage of clusterless decoding---which relates multiunit spike wave form features to position without spike sorting---giving us more information about the population spiking activity. We apply this model to data from 10 animals, focusing on SWR events to permit a direct comparison to previous work. We find that the vast majority of SWRs contain spatially coherent content. Surprisingly, while the expected high speed replay events were identified, the most common category of events expressed representations that moved at speeds consistent with real animal movement, suggesting that high speed replay is not the most common manifestation of generative activity. Our method provides a useful and powerful tool for understanding the dynamics of mental representations, and could be adapted for use in many brain areas.

@@

Once we have specified the data and dynamics model, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from the past and future spikes, to estimate the joint posterior probability of position and dynamic state (see Methods). With this, we can summarize the resulting posterior with two quantities: the mixture probability of each dynamic over time and the probability of position over time, irrespective of dynamic (Figure \ref{1}F, left and right plot respectively).

An examination of these two summaries reveals that that the model successfully captures the dynamics of the population spiking activity in Figure \ref{1}A. The stable firing of the one active neuron indicates a stationary representation, and the probability of the stationary dynamic is high at the beginning of the simulated replay. The change in the data then drives a rapid transition to the continuous dynamic, reflecting the trajectory-like spiking from the population of cells. Subsequently, as the population activity becomes spatially incoherent, the fragmented dynamic dominates for the last third. Importantly, these dynamics also inform the estimate of latent position of the animal, and the correct assignment of dynamic reduces the variance of the latent position estimate.

@@

After testing our model on a single SWR, we applied our decoding algorithm to hippocampal recordings from 10 animals performing the W-track spatial alternation task (tetrodes = [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020}). One major goal of our work was to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a relatively permissive threshold (see Methods) to ensure that we were included both the rare large events as well as the much more common small events.

@@

Previous work focusing on spatially sequential replay reported that only a minority of events contain sequential spatial content [REFS]. We therefore asked what fraction of classified events contained spatially coherent content in any form, defined as a SWR containing any times with stationary, stationary-continuous mixture, or continuous dynamics (see methods). We found that 22478 of 23382 or 96\% of classified SWRs had some spatially coherent structure. We then asked what fraction of events contained spatially incoherent content, defined as a SWR containing any times with fragmented-continuous mixture or fragmented. We found that 3353 of 23382 or 14\% of classified SWRs had some spatially incoherent structure. This was consistent across animals (Figure \ref{3}F). We then asked what percentage of classified SWR contained continuous content, like would typically be analyzed when using the Bayesian decoder. Here our findings were consistent with previous reports: in our datasets 4511 of  23382 or 19\% of classified SWRs had some continuous content (Figure \ref{3}G). Thus, focusing on only continuous content misses a large fraction of events where there is evidence for spatially coherent but not continuous content.

We repeated this analysis with a higher threshold of 0.95 to ensure that our result was not dependent on this threshold. We found that, while this slightly reduced the total fraction of classified ripples (19180 / 25845 or 74\% SWRs), an even higher fraction of the classified ripples (19020 / 19180 or 99\% classified) had spatially continuous content. Classified SWRs containing spatially incoherent content similarly were a small fraction of the classified SWRs (spatially incoherent: 484 / 19180 or 3\%).

Our model is specified in the context of a marked point process associated with different dynamics. This specification allows us to not only decode events, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so we can compute the cumulative spatial extent of the 95\% highest posterior density (HPD) values of the posterior linear position. Larger values of the cumulative HPD indicate the model is less certain about position, because the HPD values are spread out over more of the track at a given time point, whereas lower values indicate that the model is more certain about the estimate of position, because the HPD values are more concentrated and cover less of the track. Thus, the HPD provides a complementary measures of spatial coherence, and evaluating it allows us to verify that the events we defined as spatially coherent correspond to events where there is high certainty around the position estimates.

@@

Our results indicate that most events with coherent spatial content are not continuous; that is, trajectories that move many times the running speed of the animal. We therefore examined these events in more details. We first noted that most of the SWRs (16454 of 23382 classified SWRs or 70\%) were well described by a single dynamic. Surprisingly, the most common SWRs were categorized exclusively as stationary-continuous mixtures (9860 of 23382 classified SWRs or 42\%, Figure \ref{5}A). These events contain representations that move at rates consistent with real-world movement (Fig. 1G), but were slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 72 ms vs. continuous 95 ms, Figure \ref{5}B). Nonetheless, both these slow-moving events and continuous events represented locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 51 cm vs. continuous 42 cm, Figure \ref{5}C). This indicates that the trajectories of these SWRs, like ones that have faster continuous dynamics, do not represent the animal's actual position.

The next most common category is those SWRs that were exclusively stationary (5149 of 23382 or 22\%). Unlike the stationary-continuous mixtures, most of these events represented a location close to the the animal's actual position (Figure \ref{5}C). There were, however, a number of stationary events  that represented positions far from the animal. We set a threshold of 30 cm distance from the animal to identify "non-local stationary trajectories" and found these trajectories in 1585 of 23382 (~7\%) of classified SWRs. About half of these (731 classified SWRs) were stationary throughout the duration of the SWR. These non-local stationary trajectories most commonly represented reward wells or choice points (Figure \ref{5}D), consistent with \cite{JaiDistincthippocampalcorticalmemory2017}.

Finally, we examined the overall activity levels across dynamics and found that all the dynamics had similar average multiunit spiking rates by tetrode (Figure \ref{5}E). This indicates that all dynamics, including the slower dynamics such as stationary and stationary-continuous mixtures and importantly not just the continuous dynamics, were driven by sustained spiking information.

@@

@@

adjudicates/clarifies results of Pfiefer and Foster and Stella et al. Brownian motion but couldn't tell the speeds. Pfiefer and Foster didn't know the percentage of events that acted this way

Why our results are different from the Bayesian decoder
seek to describe as many SWR events, clusterless decoding, more permissive model with different dynamics, explicit descriptive model instead of implicit noisy model.

relationship to Johnson and Redish 2008, Maboudi, model represents middle ground between HMM and typical Bayesian decoder. allows SWR to quickly be identified because of the discrete states. Can work in 2D. Can precisely identify timing of dynamics because of acausal decoding

Can extend to incorporate other previously experienced environments. Add prior on dynamics of speed

@@

\subsection*{Simulated Data}
Encoding data for Figure \ref{1} and Figure \ref{fig:Figure1-Figure supplement 1} was generated by simulating 15 traversals of a 180 cm linear track. The spiking of 19 neurons was simulated using an inhomogeneous Poisson process where the instantaneous firing rate of each neuron changes according to a Gaussian place field. The Gaussian place fields had a variance of 36 cm and were spaced at 10 cm intervals over the 180 cm track. The decoding data for Figure \ref{1}G was generated by simulating 20,000 linear traversals of the 180 cm track, each with a unique constant speed, starting at 0.5 cm/s and increasing by 0.5 cm/s up to 10,000 cm/s. Each simulated neuron "spiked" when the traversal passed through the peak location of its place field.

\subsection*{Recording Locations and Techniques}
Ten male Long Evans rats (500-700 g, 4-9 months old) were trained on a W-track spatial alternation task. 9 rats contributed to previous studies \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, Kayhippocampalnetworkspatial2016, CarrTransientSlowGamma2012}. Neural activity was recorded in CA1, CA2, CA3, MEC, Subiculum, and DG depending on the rat. We only used hippocampal tetrodes (CA1, CA2, CA3) in this study.

\subsection*{Behavioral Task}
All animals performed a W-track spatial alternation task, which is described in detail in \cite{KarlssonAwakereplayremote2009}. In brief, each day, animals alternated between 20 minute rest epochs in a rest box and 15 minutes run epochs in the W-track. On the W-track, animals alternated between two types of runs: inbound and outbound. On inbound runs, the animal had to run from an outer arm to the center arm. On outbound runs, the animals had to run from the center arm to the least recently visited outer arm. Animals were rewarded for performing the task correctly. Only run epochs with at least 9 hippocampal principal cells that fired at least 100 spikes were included in the analysis.

\subsection*{Position of the Animal and Linearization}
The animal's 2D position was estimated from digital video (30 Hz) of two infrared diodes placed on the head stage preamplifiers using a semi-automated analysis. In order to decrease the time it takes to run the model, the 2D position of the animal was converted into a 1D position. This is done by first defining a 2D graph representation of the track, where edges correspond to segments of the W-track and nodes represent intersection points between those segments. Then, based on the algorithm in \cite{NewsonHiddenMarkovmap2009}, we use a Hidden Markov Model (HMM) to classify the most likely track segment the animal is on. Using the HMM takes into account the time dependent nature of the data and helps prevents sudden jumps from one track segment to another, which is particularly important near intersections. The observation model of the HMM is Gaussian and it models the likelihood of being on a track segment as the Gaussian distance to that segment with a 5 cm standard deviation. The state transition model is an empirically estimated state transition that changes with each time point that tries to ensure that the Euclidean distance between successive position estimates is similar to the shortest path distance along the graph between successive position estimates. A slight bias of 0.1 is given to the diagonal to encourage staying on the same track segment. The most likely track segment the animal is on is computed using the Viterbi algorithm. After finding the track segment that corresponds to each 2D position, the 2D position is projected onto the track segment. This allows us to define a distance from the center well in terms of shortest path length on the track, where 0 cm represents the center well position. The linear distance can then be converted into a linear position by assigning each track segment a position in 1D space. 15 cm gaps were placed between the center arm, left arm, and right arms in 1D space to prevent any smoothing done in the model from bleeding over inappropriately. The code used for linearization can be found at \url{https://github.com/Eden-Kramer-Lab/loren_frank_data_processing}.

\subsection*{Sorted Spikes, Multiunit Spikes and Waveform Features}
To obtain the neural data used for decoding, hippocampus electrical potentials were recorded at 30 kHz, referenced to a corpus callosum tetrode, and then digitally filtered between 600 Hz and 6 kHz. Events were considered threshold crossing events if any one of the four tetrode wires exceeded a 60 mV threshold. These multiunit threshold crossing events and the associated peak amplitudes of the tetrodes wires---defined as the electrical potential value at the time of the maximum amplitude at the threshold crossing---were used as data for the clusterless version of the model.

For decoding using sorted spikes, the multiunit events were processed further to assign events to putative single cells. Putative single cells were manually identified based on the clustering of three waveform features within a day: peak amplitude, principal components, and spike width. Only putative hippocampal pyramidal cells---identified based on spike width and average firing rate---were included in the analysis.

\subsection*{SWR Detection}
Sharp wave ripples were detected using the same method as in \cite{Kayhippocampalnetworkspatial2016}. Each CA1 LFP was obtained by downsampling the original 30 kHz electrical potential to 1.5 kHz, bandpass filtering between 0.5 Hz and 400 Hz. This was further bandpass filtered between 150-250 Hz for ripple band frequencies, squared and then summed across tetrodes---forming a single population trace over time. This trace was smoothed with a Gaussian with a 4 ms standard deviation and the square root of this trace was taken to get an estimate of the population ripple band power. Candidate SWR times were found by z-scoring the population power trace and finding times when the z-score exceeded 2 standard deviations for a minimum of 15 ms and the speed of the animal was less than 4 cm/s. The SWR times were then defined as times when the z-score was greater than the mean and contained the candidate SWR times. The code used for ripple detection can be found at \url{https://github.com/Eden-Kramer-Lab/ripple_detection}.

\subsection*{The Model}
Let $x_{k}$ be a continuous latent variable that corresponds to the position represented by the population of cells at time $t_k$ and $I_{k}$ be a discrete latent variable that is an indicator for the dynamics we wish to characterize: stationary, continuous, and fragmented. The goal of the model is to simultaneously estimate the posterior probability of position and dynamics $p(x_k, I_k \mid O_{1:T})$, where $O_{1:T}$ corresponds to the observed spiking data from time 1 to time $T$. The observed data can be either spike trains $\Delta N_{1:T}^{(1:C)}$ from $C$ putative cells when decoding with sorted spikes or multiunit spikes $\Delta N_{1:T}^{(1:E)}$ and their associated wave form features $\Vec{m}$ from each tetrode $E$ when decoding with clusterless spikes.

We have previously shown \cite{DenovellisCharacterizinghippocampalreplay2019} that the posterior probability $p(x_k, I_k \mid O_{1:T})$ can be estimated by applying the following recursive causal filter equation, starting with initial conditions $p(x_{0}, I_{0} \mid O_{0})$ and iterating to time T:

$$
p(x_{k}, I_{k} \mid O_{1:k}) \propto
p(O_{k}  \mid x_{k}, I_{k}) \sum_{I_{k-1}} \int p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) 
Pr(I_{k} \mid I_{k-1}) p(x_{k-1}, I_{k-1} \mid O_{1:k-1}) dx_{k-1}
$$
and then applying the acausal smoother equation, starting from the last estimate of the casual filter $p(x_{T}, I_{T} \mid O_{1:k})$ and recursively iterating backwards to time 1:
$$
p(x_{k}, I_{k} \mid O_{1:T}) =
p(x_{k}, I_{k} \mid O_{1:k})
\sum_{I_{k+1}} \int \frac{p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k})
Pr(I_{k+1} \mid I_{k})}{p(x_{k+1}, I_{k+1} \mid O_{1:k})}
p(x_{k+1}, I_{k+1} \mid O_{1:T}) dx_{k+1}
$$
where:
$$
p(x_{k+1}, I_{k+1} \mid O_{1:k}) =
\sum_{I_{k}} \int p(x_{k+1} \mid x_{k}, I_{k+1}, I_{k}) Pr(I_{k+1} \mid I_{k})
p(x_{k}, I_{k} \mid O_{1:k}) dx_{k}
$$

@@

For the dynamics transition matrix $Pr(I_{k} \mid I_{k-1})$, which defines how likely the dynamic is to change to another dynamic versus persist in the same dynamic, we set it to be:

@@

For the dynamics movement model $p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1})$, which defines how likely the position $x_{k}$ is to change given the previous position $x_{k-1}$ and current $I_{k}$ and past dynamics $I_{k-1}$, we set it to be:

\begin{equation*}
  p(x_{k} \mid x_{k-1}, I_{k}, I_{k-1}) = 
  \begin{blockarray}{*{3}{c} l}
    \begin{block}{*{3}{>{$\footnotesize}c<{$}} l}
      $I_{k} = Stationary$ & $I_{k} = Continuous$ & $I_{k} = Fragmented$ \\
    \end{block}
    \begin{block}{[*{3}{c}]>{$\footnotesize}l<{$}}
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0) & \mathcal{U}(\min x, \max x) \bigstrut[t]& $I_{k-1} = Stationary$ \\
      \delta(x_{k-1}) & \mathcal{N}(x_{k-1}, 6.0)  & \mathcal{U}(\min x, \max x) & $I_{k-1} = Continuous$ \\
      \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & \mathcal{U}(\min x, \max x) & $I_{k-1} = Fragmented$ \\
    \end{block}
  \end{blockarray}
\end{equation*}
where $\delta(x_{k-1})$ is an identity transition matrix where position cannot change from the previous time step, $\mathcal{N}(x_{k-1}, 6.0)$ is a random walk from the previous position with variance 6.0, and $\mathcal{U}(\min x, \max x)$ is a uniform transition that allows transitions to any possible position. As discussed in the Results, this says that when persisting in the same dynamic, the stationary, continuous, and fragmented dynamics are defined by the identity transition, the random walk, and the uniform transition, respectively. When transitioning to or from the fragmented dynamic, we assume we do not have any information about the position, so the transition is uniform. Finally, when the transition is from the stationary to continuous, we assume the position is spatially close where it was previously, so we use a random walk. When the transition is from continuous to stationary, we assume that the position is no longer changing, so we use the identity transition.

Lastly, we evaluate the likelihood of the observations $p(O_{k} \mid x_{k}, I_{k})$ based on an encoding model fit during the encoding period. We assume the likelihood is the same for each dynamic $I_k$, so we only need to estimate $p(O_{k} \mid x_{k})$. It has been shown \cite{ZhangInterpretingNeuronalPopulation1998, BrownStatisticalParadigmNeural1998} that the likelihood with sorted spikes can be computed as:
$$
p(O_{k} \mid x_{k}) = p(\Delta N_{k}^{(1:C)} \mid x_{k}) \propto
\prod^{C}_{i=1} [\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]^{N_{t_{k}}^{i}} \\
    \exp[-\lambda_{i}(t_{k} \mid x_{k})\Delta_{k}]
$$
where $N_{t_{k}}^{i}$ represents a spike at time $t_k$ from cell $i$, $\Delta_k$ is the time bin size, and $\lambda_{i}(t_{k} \mid x_{k})$ is the instantaneous firing rate of cell $i$ given position $x_{k}$. $\lambda_{i}(t_{k} \mid x_{k})$ is the "place field" of the cell, which can be estimated via generalized linear model during the encoding period (which we explain further in the Model Implementation section).
